Article Title

### 2014-01-27 plos.tex
minor

\begin{document}
% Title must be 150 characters or less
\begin{flushleft}
{\Large
Improved estimation and interpretation of correlations in neural circuits
}
% Insert Author names, affiliations and corresponding author email.
\\
Dimitri Yatsenko,$^{1,\ast}$, 
Kre\v{s}imir Josi\'{c}$^{2}$,
Alexander S.~Ecker$^{1,3,4}$,
Emmanouil Froudarakis$^{1}$,
R.~James Cotton$^{1}$,
Andreas S.~Tolias$^{1,5}$
\\
\bf{1} Department of Neuroscience, Baylor College of Medicine, Houston, TX, USA
\\
\bf{2} Department of Mathematics and Department of Biology and Biochemistry, University of Houston, Houston, TX, USA
\\
\bf{3}  Werner Reichardt Center for Integrative Neuroscience and Institute for Theoretical Physics, University of T\"ubingen, Germany
\\
\bf{4} Bernstein Center for Computational Neuroscience, T\"ubingen, Germany
\\
\bf{5} Department of Computational and Applied Mathematics, Rice University, Houston, TX, USA

$\ast$ E-mail: yatsenko@cns.bcm.edu
\end{flushleft}

@@

In the example site (Fig.~\ref{fig:05}), the sparse component had 82.2\% sparsity (or conversely, 17.8\% connectivity: $connectivity=1-sparsity$) with average node degree of 52.9 (\figref{05}{G}). The averge node degree, \ie the average number of interactions linking each neuron, is related to connectivity as $degree = connectivity\cdot(p-1)$. The low-rank component had rank 15, expressing 15 inferred latent units. The number of latent units and the average node degrees genearlly increased with population size, although these values varied widely between the datasets \figref{06}{A and B}. There was an inverse relationship between the number of latent units and the average node degree (\figref{06}{C}): Several sites, despite their large population sizes, were driven by latent units and had few pairwise interactions. This variability may be explained by differences in brain states and recording quality and needs further investigation.

The average partial correlations calculated from these estimates according to Eq.~\ref{eq:partial} at all 31 sites were about 5 times lower than the average sample correlations (\figref{06}{D}). This signifies that correlations between neurons build up from large numbers of smaller partial correlations or interactions with other neurons. Furthermore, the avearge partial correlations were less variable: the coefficient of variation of the average sample correlations across sites was 0.41 whereas that of the average partial correlations was only 0.35.

While the sample correlations were mostly positive, the sparse component of the partial correlations (\sq{interactions}) had a high fraction (43.3\%) of negative (\sq{inhibitory}) values (\figref{05}{F}). This fraction was fairly consistent across multiple sites but the fraction of negative interactions increased with the inferred connectivity \figref{06}{F}.  This increase signifies that negative interactions are inferred only after a sufficient density of positive interactions have been uncovered.

@@

We compared the graphs of interactions inferred by $C_{\sf sparse+latent}$ to the graph obtained by thresholding the sample correlations.   Thresholded sample correlations have been used in several studies to infer pairwise interactions \cite{Golshani:2009, Feldt:2011, Malmersjo:2013}.  We set the threshold to match the connectivity of the $C_{\sf sparse+latent}$ graph.  The networks revealed by the two methods differed substantially. In the example site with 17.8\% connectivity inferred by $C_{\sf sparse+latent}$, only 41.7\% of connections coincided with above-threshold sample correlations (\figref{05}{F, H, and I}). In particular, most of the inferred negative interactions corresponded to low sample correlations (\figref{05}{F}) where high correlations were expected given the rest of the correlation matrix.   This low overlap between the two types of graphs was consistent across across the datasets even though  their connectivity varied \figref{06}{E}.

@@

\section*{Discussion}
\paragraph{Functional connectivity as a network of pairwise interactions}
Functional connectivity is often represented as a graph of pairwise interactions. The goal of many studies of functional connectivity was to estimate  anatomical connectivity from  observed multineuronal spiking activity.  For example, characteristic peaks and troughs in the pairwise cross-correlograms of recorded spike trains contain statistical signatures of directional monosynaptic connections and shared synaptic inputs \cite{Gerstein:1964, Perkel:1967, Moore:1970, Alonso:1998, Denman:2013}.  Such signatures are ambiguous as they can arise from network effects other than direct synaptic connections \cite{Aertsen:1989}.  With simultaneous recordings from more neurons, ambiguities can be resolved by inferring the conditional dependencies between pairs of neurons.  Direct causal interactions between neurons produce statistical dependency between them even after conditioning on the state of the remainder of the network \Kcomment{and the input?}. Therefore, conditionally independent neurons can be inferred to lack a direct causal influence.

Conditional dependencies can be inferred by fitting a probabilistic model of the joint population activity. For example, generalized linear models (GLMs) have been constructed to  include biophysically plausible synaptic integration and membrane kinetics, and individual neurons' stimulus drive~\cite{Pillow:2008}.  Maximum entropy models are similarly constrained by \Kcomment{stimulus drive and} observed pairwise correlations \cite{Schneidman:2006, Tkacik:2006, Yu:2008, Tang:2008, Shlens:2009}.  
\Kcomment{Assuming that the population response follows a} multivariate normal distribution, the conditional dependencies between pairs of neurons are expressed by the partial correlations between them.   Each probabilistic model, fitted to the same data may reveal a different network of \sq{interactions},  \ie conditional dependencies between pairs of cells.

It is not yet clear which approach provides the best correspondence with anatomical connectivity. Little experimental evidence is available to answer this question.  The connectivity graphs inferred by various statistical methods are commonly reported without examining their relation to \Kcomment{anatomy}.  
Topological properties of such graphs have been interpreted as principles of circuit organization (\emph{e.g.} small-world organization) \cite{Feldt:2011, Yu:2008, Malmersjo:2013}.  However, the topological properties of functional connectivity graphs can depend on the method of inference \cite{Zalesky:2012}. Until a physiological interpretation of functional connectivity is established, the validity of such analyses remains in question.

Inference of the conditional dependencies also depends on the completeness of the recorded population:  To properly condition the pairwise interactions on the activity of the other neurons, we must record from all neurons with which the pair interacts. Unobserved portions of the circuit can introduce false conditional dependencies between observed neurons. For this reason, statistical models of population activity have been most successfully applied to \emph{in vitro} preparations of the retina or cell cultures where high-quality recordings from the complete populations were available \cite{Pillow:2008}. In cortical tissue, electrode arrays record from a small fraction of cells in a given volume, limiting the validity of inference of the pairwise conditional dependencies. This may be the reason that partial correlations have not previously been used to describe  functional connectivity in cortical populations.

Two-photon imaging of population calcium signals presents unique advantages for the estimation of functional connectivity.  While the temporal resolution of calcium signals is limited by calcium dye kinetics, fast imaging techniques combined with spike inference algorithms provide millisecond-scale temporal resolution of single action potentials \cite{Grewe:2010}. However, such high temporal precision comes at the cost of the accuracy of inferred spike rates.  Better accuracy is achieved when calcium signals are analyzed on scales of tens of milliseconds \cite{Cotton:2013}.  The major advantage of calcium imaging is its ability to characterize the spatial arrangement and types of recorded cells.  Recently, advanced imaging techniques have allowed for recordings from nearly every cell in a volume of cortical tissue  \emph{in vivo} \cite{Katona:2012, Cotton:2013} and even from entire nervous systems \cite{Leung:2013, Ahrens:2013}.  These techniques may provide more incisive measurements of functional connectivity than classical electrophysiological recordings.

@@

Coactivation patterns and pairwise connectivity are not mutually exclusive since assemblies arise from patterns of synaptic connectivity.  However, an analysis of coactivation shifts the focus from detailed interactions to  collective behavior.  
In our study, the functional connectivity through modes of coactivations was represented by the factor analysis estimator $C_{\sf factor}$.

@@

\paragraph{Addressing ill-posedness}
Inferring the conditional dependencies between variables in a probabilistic model is an ill-posed problem: small variations in the data produce large errors in the inferred network of dependencies.  \Kcomment{References for the previous statement?} The problem becomes worse as the number of  recorded neurons increases until such models lose their statistical validity \cite{Roudi:2009}.  As  techniques improved to allow for recording from larger neuronal populations, experimental neuroscientists addressed this problem by extending the recording durations to keep sampling noise in check and verified that existing models are not overfitted and that the stationarity assumption is observed\cite{Tkacik:2013}. \Kcomment{The stationarity assumption seems a bit orthogonal to the rest discussed here.}  However,  ambitious projects such as the BRAIN initiative  \cite{Alivisatos:2013}, aim to record from significantly larger populations. Simply increasing recording duration will not be sufficient, and the problem must be addressed using regularized estimators. Regularization biases the solution toward a small subspace to counteract the effect of  sampling noise in the empirical data. However, biasing the solution to an inappropriate subspace does not allow significant estimation improvement and hampers interpretation.

@@

The results of model selection must be interpreted with caution.  As we demonstrated by simulation, with limited data an incorrect, but simpler (lower dimension) model can produce a better estimate than a more complex model with correct representation of dependencies.   Therefore, showing that a more constrained model has better cross-validated performance than a more complex model does not support the conclusion that it reveals a better representation of dependencies in the data.  This caveat is related to \emph{Stein's Paradox} \cite{Efron:1977}: The biasing of an estimate toward an arbitrary target is guaranteed to produce \emph{some} improvement over the best unbiased estimate.

@@

Moreover, this estimation approach provides a graphical representation of the dependencies in the data that can be used to formulate and test hypotheses about the structure of connectivity in the microcircuit. Importantly, the inferred functional interactions were substantially different from those measured by considering the network of the most significant sample correlations \figref{05}{F, H, and I}.  For example, the $C_{\sf sparse+latent}$ estimator reveals a large number of negative interactions that were not present in the sample correlation matrix (\figref{05}{F}) that may reflect inhibitory circuitry.  Although the network of functional interactions inferred by the optimal estimator, $C_{\sf sparse+latent}$, will not be identical to the anatomical connectivity it may accurately reflect average connectivity profiles. Future experiments utilizing molecular markers for different cell types, and follow-up multiple whole-cell in vitro experiments \cite{Hofer:2012, Ko:2013} \Kcomment{ref} will directly compare the inferred functional connectivity graphs to the underlying anatomical circuitry. Finally, the latent units inferred by the estimator can be analyzed for their physiological interpretation. For example, these latent units may be modulated under different brain states (e.g. slow-wave sleep, attention) and stimulus conditions (e.g. certain types of stimuli may engage feedback connections)

@@

\section*{Acknowledgments}
We thank Genevera Allen for a helpful discussion, and Eftychios Pnevmatikakis for helpful suggestions and feedback.

% The bibtex filename
\bibliography{references.bib}

>>>>

\begin{document}
% Title must be 150 characters or less
\begin{flushleft}
{\Large
Improved estimation and interpretation of correlations in neural circuits
}
% Insert Author names, affiliations and corresponding author email.
\\
Dimitri Yatsenko,$^{1}$, 
Kre\v{s}imir Josi\'{c}$^{2}$,
Alexander S.~Ecker$^{1,3,4}$,
Emmanouil Froudarakis$^{1}$,
R.~James Cotton$^{1}$,
Andreas S.~Tolias$^{1,5,\ast}$
\\
\bf{1} Department of Neuroscience, Baylor College of Medicine, Houston, TX, USA
\\
\bf{2} Department of Mathematics and Department of Biology and Biochemistry, University of Houston, Houston, TX, USA
\\
\bf{3}  Werner Reichardt Center for Integrative Neuroscience and Institute for Theoretical Physics, University of T\"ubingen, Germany
\\
\bf{4} Bernstein Center for Computational Neuroscience, T\"ubingen, Germany
\\
\bf{5} Department of Computational and Applied Mathematics, Rice University, Houston, TX, USA

$\ast$ E-mail: atolias@cns.bcm.edu
\end{flushleft}

@@

In the example site (Fig.~\ref{fig:05}), the sparse component had 82.2\% sparsity (or conversely, 17.8\% connectivity: $connectivity=1-sparsity$) with average node degree of 52.9 (\figref{05}{G}). The averge node degree, \ie the average number of interactions linking each neuron, is related to connectivity as $degree = connectivity\cdot(p-1)$, where $p$ is the number of neurons. The low-rank component had rank 15, expressing 15 inferred latent units. The number of latent units and the average node degrees generally increased with population size, although these values varied widely between the datasets (\figref{06}{A and B}). There was an inverse relationship between the number of latent units and the average node degree (\figref{06}{C}): Several sites, despite their large population sizes, were driven by latent units and had few pairwise interactions. This variability may be explained by differences in brain states and recording quality and needs further investigation.

The average partial correlations calculated from these estimates according to Eq.~\ref{eq:partial} at all 31 sites were about 5 times lower than the average sample correlations (\figref{06}{D}). This suggests that correlations between neurons build up from large numbers of small interactions with other neurons. Furthermore, the avearge partial correlations were less variable: the coefficient of variation of the average sample correlations across sites was 0.41 whereas that of the average partial correlations was 0.35.

While the sample correlations were mostly positive, the sparse component of the partial correlations (\sq{interactions}) had a high fraction (43.3\%) of negative values (\figref{05}{F}). This fraction was consistent across multiple sites but the fraction of negative interactions increased with the inferred connectivity \figref{06}{F}.  This increase suggests that negative interactions are inferred only after a sufficient density of positive interactions have been uncovered.

@@

Thresholded sample correlations have been used in several studies to infer pairwise interactions \cite{Golshani:2009, Feldt:2011, Malmersjo:2013}.  We therefore compared the interactions inferred by $C_{\sf sparse+latent}$ to those obtained by thresholding the sample correlations with the threshold set to match the connectivity of the $C_{\sf sparse+latent}$ graph.  The networks revealed by the two methods differed substantially. In the example site with 17.8\% connectivity inferred by $C_{\sf sparse+latent}$, only 41.7\% of connections coincided with above-threshold sample correlations (\figref{05}{F, H, and I}). In particular, most of the inferred negative interactions corresponded to low sample correlations (\figref{05}{F}) where high correlations were expected given the rest of the correlation matrix.  This low overlap between the two types of graphs was consistent across the datasets \figref{06}{E}. 
t

@@

\section*{Discussion}
\paragraph{Functional connectivity as a network of pairwise interactions}
Functional connectivity is often represented as a graph of pairwise interactions. The goal of many studies of functional connectivity was to estimate  anatomical connectivity from  observed multineuronal spiking activity.  For example, characteristic peaks and troughs in the pairwise cross-correlograms of recorded spike trains contain statistical signatures of directional monosynaptic connections and shared synaptic inputs \cite{Gerstein:1964, Perkel:1967, Moore:1970, Alonso:1998, Denman:2013}.  Such signatures are ambiguous as they can arise from network effects other than direct synaptic connections \cite{Aertsen:1989}.  With simultaneous recordings from more neurons, ambiguities can be resolved by inferring the conditional dependencies between pairs of neurons.  Direct causal interactions between neurons produce statistical dependency between them even after conditioning on the state of the remainder of the network and external input. Therefore, conditionally independent neurons can be inferred to lack a direct causal influence.

Conditional dependencies can be inferred by fitting a probabilistic model of the joint population activity. For example, generalized linear models (GLMs) have been constructed to  include biophysically plausible synaptic integration and membrane kinetics, and individual neurons' stimulus drive~\cite{Pillow:2008}.  Maximum entropy models are similarly constrained by observed pairwise correlations \cite{Schneidman:2006, Tkacik:2006, Yu:2008, Tang:2008, Shlens:2009} .  
Under the multivariate normal distribution, the conditional dependencies between pairs of neurons are expressed by the partial correlations between them.   Each probabilistic model, fitted to the same data may reveal a different network of \sq{interactions},  \ie conditional dependencies between pairs of cells.

It is not yet clear which approach provides the best correspondence with anatomical connectivity. Little experimental evidence is available to answer this question.  The connectivity graphs inferred by various statistical methods are commonly reported as abstract representations of the functional connectivity without examining their relation to anatomical implications.  
Topological properties of such graphs have been interpreted as principles of circuit organization (\emph{e.g.} small-world organization) \cite{Feldt:2011, Yu:2008, Malmersjo:2013}.  However, the topological properties of functional connectivity graphs vary with the method of their inference \cite{Zalesky:2012}. Until a physiological interpretation of functional connectivity is established, the validity of such analyses remains in question.

Inference of the conditional dependencies also depends on the completeness of the recorded population:  To properly condition the pairwise interactions on the activity of the other neurons, we must record from all neurons with which the pair interacts. Unobserved portions of the circuit can introduce false conditional dependencies between observed neurons. For this reason, the most statistical models of population activity have been most successfully applied to \emph{in vitro} preparations of the retina or cell cultures where high-quality recordings from the complete populations were available \cite{Pillow:2008}. In cortical tissue, electrode arrays record from a small fraction of cells in a given volume, limiting the validity of the inference of the pairwise conditional dependencies. This may be the reason that partial correlations have not previously been used to describe  functional in cortical populations.

Two-photon imaging of population calcium signals presents unique advantages for the estimation of functional connectivity.  While the temporal resolution of calcium signals is limited by calcium dye kinetics, fast imaging techniques combined with spike inference algorithms provide millisecond-scale temporal resolution of single action potentials \cite{Grewe:2010}. However, such high temporal precision comes at the cost of the accuracy of inferred spike rates.  Better accuracy is achieved when calcium signals are analyzed on scales of tens of milliseconds \cite{Cotton:2013}.  The major advantage of calcium imaging is its ability to characterize the spatial arrangement and types of recorded cells.  Recently, advanced imaging techniques have allowed for recordings from nearly every cell in a volume of cortical tissue in vivo \emph{in vivo} \cite{Katona:2012, Cotton:2013} and even from entire nervous systems \cite{Leung:2013, Ahrens:2013}.  These techniques may provide more incisive measurements of functional connectivity than classical electrophysiological recordings.

@@

Coactivation patterns and pairwise connectivity are not mutually exclusive since assemblies arise from patterns of synaptic connectivity, but an analysis of coactivation shifts the focus from detailed interactions to  collective behavior.  
In our study, the analysis of functional connectivity through modes of coactivations was represented by the factor analysis estimator $C_{\sf factor}$.

@@

\paragraph{Addressing ill-posedness}
Inferring the conditional dependencies between variables in a probabilistic model is an ill-posed problem: small variations in the data produce large errors in the inferred network of dependencies.  The problem becomes worse as the number of  recorded neurons increases until such models lose their statistical validity \cite{Roudi:2009}.  As  techniques improved to allow for recording from larger neuronal populations, experimental neuroscientists addressed this problem by extending the recording durations to keep sampling noise in check and verified that existing models are not overfitted and that the stationarity assumption is observed\cite{Tkacik:2013}.  However,  ambitious projects such as the BRAIN initiative  \cite{Alivisatos:2013}, aim to record from significantly larger populations. Simply increasing recording duration will not be sufficient, and the problem must be addressed using regularized estimators. Regularization biases the solution toward a smaller subspace to counteract the effect of the sampling noise in the empirical data. However, biasing the solution to an inappropriate subspace does not allow significant estimation improvement and hampers interpretation.

@@

The results of model selection must be interpreted with caution.  As we demonstrated by simulation, a wrong simple model can produce a better estimate than a more complex model with correct representation of dependencies.   Therefore, showing that a more constrained model has better cross-validated performance than a more complex model does not support the conclusion that it reveals a better representation of dependencies in the data.  This caveat is related to \emph{Stein's Paradox} \cite{Efron:1977}: in which the biasing of an estimate toward an arbitrary target is guaranteed to produce \emph{some} improvement over the best unbiased estimate.

@@

Moreover, this estimation approach provides a graphical representation of the dependencies in the data that can be used to formulate and test hypotheses about the structure of connectivity in the microcircuit. Importantly, the inferred functional interactions were substantially different from those measured by considering the network of the most significant sample correlations \figref{05}{F, H, and I}.  For example, the $C_{\sf sparse+latent}$ estimator reveals a large number of negative interactions that were not present in the sample correlation matrix (\figref{05}{F}) that may reflect inhibitory circuitry.  Although the network of functional interactions inferred by the optimal estimator, $C_{\sf sparse+latent}$, will not be identical to the anatomical connectivity it may accurately reflect average connectivity profiles. Future experiments utilizing molecular markers for different cell types, and follow-up multiple whole-cell in vitro experiments \cite{Hofer:2011, Ko:2013} will directly compare the inferred functional connectivity graphs to the underlying anatomical circuitry. Finally, the latent units inferred by the estimator can be analyzed for their physiological interpretation. For example, these latent units may be modulated under different brain states (e.g. slow-wave sleep, attention) and stimulus conditions (e.g. certain types of stimuli may engage feedback connections).

@@

\section*{Acknowledgments}
We thank Genevera Allen for a helpful discussion, and Eftychios Pnevmatikakis for helpful suggestions and feedback on the manuscript.  This work was supported by grants NEI R01-EY018847, NEI P30-EY002520-33, NEI T32-EY07001-37, NIDA R01DA028525, the NIH-Pioneer award DP1-OD008301 to AST, the McKnight Scholar Award to AST; the Arnold & Beckman Foundation Young Investigator Award to AST; and NSF grants DMS-0817649, DMS-1122094 to KJ.
% The bibtex filename
\bibliography{references.bib}
